\section{Design}
\label{sec:design}

\sys{} is a programmable OS-level control plane for agent harnesses. It
compiles behavioral policies into labeled information-flow rules at OS
kernel hooks, observing and enforcing agent behavior and returning semantic
feedback when a rule is matched. Figure~\ref{fig:architecture} summarizes
the pipeline.

\subsection{Threat Model and Assumptions}

\sys{} targets a \emph{cooperative but unreliable} agent: the agent is
trusted to declare its task intent and policy constraints honestly, but
not trusted to follow them reliably during execution. The agent may make
mistakes because of long-horizon planning errors, prompt injection,
shell-script side effects, or opaque tool behavior. \sys{} does not
target a malicious agent that deliberately attacks the kernel, exploits
the loader, or attempts to evade policy through adversarial
obfuscation.

The trusted computing base consists of the eBPF programs, BPF maps, DSL
compiler, runner, feedback formatter, and higher-authority policies. The
agent's process tree, including tools, shells, subprocesses, and
sub-agents, is treated as the monitored domain. Policy authority and
conflict resolution are handled by the layered composition model described
below.

Out of scope are kernel compromise, root or \texttt{CAP\_BPF}
compromise, deliberate side channels, and semantic content checks that
are not observable at process, file, or network edges. These assumptions
match the harness setting we target: cooperative coding agents that need
deterministic runtime guardrails for their own system-level behavior.

\subsection{Design Goals}

The empirical study in \S\ref{sec:empirical} gives four requirements:
agent-provided context, OS-boundary enforcement, cross-event state, and
semantic feedback. \sys{} adopts these as system goals and adds one
authority goal: agent-authored rules may specialize the session policy,
but they must not weaken operator or platform constraints.

\subsection{Policy Lifecycle}

\sys{} separates \emph{who supplies context} from \emph{who enforces the
resulting rule}. Higher-authority policies, such as platform or operator
constraints, are supplied before the agent starts and define the
non-negotiable boundary of the session. Project policies come from
instruction files and repository configuration. Session policies are
agent-authored: before a protected task begins, the agent reads the
project instructions, resolves project context such as test commands or
protected paths, incorporates task context such as user approval or task
scope, and writes the resulting DSL policy.

This lifecycle is the mechanism behind the ``agent as control plane''
claim. A directive such as ``run the full test suite before committing''
does not name a concrete command. A static OS policy cannot know whether
the project uses \texttt{go test ./...}, \texttt{pytest}, or
\texttt{pnpm test}. The agent can inspect the repository, bind the
directive to the current project, and install a rule such as
\texttt{block exec "git" "commit" unless after exec "go" since write
"**/*.go"}. Task-scoped directives are handled similarly: a rule involving
``unless explicitly requested'' is generated from the current user request
or approval state, not from the repository alone.

Policy adaptation is explicit. When the task changes or the agent resolves
new project context, it produces a new policy document and asks the harness
to compile and load it at the next session or command boundary. The
currently loaded policy remains the authority until the replacement has
compiled successfully. This avoids an unsafe interval in which a failed
policy update silently removes enforcement.

\begin{figure}[t]
\begin{verbatim}
actplane.yaml
  policy: |
        |
        v
Rust compiler ------> struct taint_config
        |                    |
        v                    v
 actplane run -----> eBPF loader + BPF maps
        |                    |
        v                    v
  target command <--- kernel hooks
        |
        v
.actplane/last-violation.txt
\end{verbatim}
\caption{\sys{} pipeline. The Rust compiler lowers the policy DSL into a
fixed-size binary configuration. The eBPF loader writes it into read-only
data, attaches to kernel hooks, and reports rule matches as NDJSON. The runner
maps rule identifiers back to human-readable feedback.}
\Description{ASCII diagram of policy compilation, eBPF loading, target
command execution, and feedback output.}
\label{fig:architecture}
\end{figure}

\subsection{Layered Policy Composition}

\sys{} separates a pure IFC rule from the runtime binding that makes it
active. A rule is a condition, an effect, and a human-readable reason; by
itself, it does not define inheritance or authority. A \emph{domain} is
the runtime policy boundary for a process tree: each process maps to one
domain, and each event is checked against that domain's effective policy.
Files, endpoints, and stdio are IFC objects that carry labels, not policy
domains.

A binding attaches a rule to a domain as either \emph{locked} or
\emph{default}. Locked bindings are security invariants inherited by child
domains and cannot be disabled below the binding point. Default bindings
are templates: a child domain may keep them, disable them for itself, or
replace them with stricter local policy. Thus a child domain may add
constraints, labels, gates, or narrower scope, but it cannot disable
inherited locked bindings, clear inherited labels or gates, widen scope,
or mutate the definition of an inherited locked rule.

\sys{} composes policies monotonically across authority layers:
platform/operator rules, project-maintainer rules, and agent/session
rules. Lower-authority layers may add constraints or more specific
feedback, but they cannot delete, rewrite, declassify, or weaken rules
introduced by a higher-authority layer. Exceptions to a higher-authority
rule must be encoded in that same layer; an agent-authored policy cannot
turn a platform \texttt{block} into a \texttt{notify}, nor can it
declassify labels created by the platform policy. The prototype resolves
domain bindings at compile/load boundaries; the kernel receives compact
compiled rule tables and does not parse YAML or DSL on the event path.

At compile time, the policy compiler merges all active layers into one
rule set and preserves each rule's authority level in metadata. The kernel
evaluates the merged rule set against each event. If multiple clauses match
the same event, effects are joined by severity:
\[
\texttt{kill} > \texttt{block} > \texttt{notify} > \texttt{allow}.
\]
Thus lower-authority policy can only make the final decision stricter. If
a project rule says \texttt{notify} on \texttt{git push} but a platform
rule says \texttt{block} on the same event, the operation is blocked. If
two matched rules have the same effect, \sys{} reports the
higher-authority rule first, then uses deterministic rule order as a
tiebreaker. The feedback file may include additional matched reasons, but
the enforcement decision is always the strongest joined effect.

\subsection{Policy Model}

\sys{} policies are object-level information-flow rules over three node
types: processes (identified by pid and command name), files (keyed by
path hash), and endpoints (keyed by IPv4 address and mask). The model
operates at object granularity rather than byte
granularity~\cite{kemerlis2012libdft}, trading precision for low-overhead
whole-system coverage.

\subsubsection{Labels and Propagation}

Each node carries a 64-bit label bitmask. A \emph{source} declaration
introduces labels: an exec source labels processes that execute a matching
binary, a file source labels files matching a path pattern, and an endpoint
source labels network endpoints matching an address range.
Let $L(x)$ denote the label bitmask of node $x$. Labels propagate through
fixed transfer functions:

\begin{itemize}
\item \textbf{fork}$(p \to c)$: $L(c) := L(c) \cup L(p)$.
\item \textbf{exec}$(p, f)$: $L(p) := L(p) \cup L(f) \cup S(f)$, where
$S(f)$ is the set of source labels matching $f$.
\item \textbf{read}$(p, f)$: $L(p) := L(p) \cup L(f)$.
\item \textbf{write}$(p, f)$: $L(f) := L(f) \cup L(p)$.
\item \textbf{connect}$(p, e)$: $L(e) := L(e) \cup L(p)$.
\end{itemize}

All transfer functions are monotonic: labels accumulate and are never
removed by propagation. The only mechanism that removes a label is an
explicit \emph{declassify} directive in the DSL, which clears specified
labels when the agent executes a designated gate binary (e.g., a redaction
tool). This monotonicity bounds over-tainting: a node's label set grows
at most to the union of all source labels reachable through observed flow
edges.

These transfer functions maintain a provenance invariant: if a process $p$
carries label $\ell$, then information from some source tagged with $\ell$
has reached $p$ through observed OS-level flow edges.

\subsubsection{Rules and Effects}

Unlike traditional taint analysis and IFC systems that prioritize
completeness (minimizing false negatives, tolerating high false
positives)~\cite{enck2010taintdroid,clause2007dytan}, \sys{} targets
\emph{precision}: false positives that block legitimate operations
disrupt the agent's development workflow and erode trust in the harness.
The design therefore favors object-granularity labels with explicit
declassification gates over byte-level tracking, accepting bounded
under-tainting in exchange for near-zero false positives on compliant
executions.

A rule contains one or more clauses. Each clause starts with an action verb
(\texttt{notify}, \texttt{block}, or \texttt{kill}) followed by an
operation (\texttt{exec}, \texttt{read}, \texttt{write}, \texttt{unlink},
\texttt{connect}, or \texttt{open}), a target pattern, optional positional
arguments, a Boolean expression over required and forbidden labels, and an
optional condition (target allow-list, lineage gate, or temporal after-gate).

\begin{figure}[t]
\begin{verbatim}
source AGENT = exec "claude"
source SECRET = file "**/.env"

# Per-event: from CoplayDev/unity-mcp
rule no-push-main:
  kill exec "git" "push" "main" if AGENT
  because "Do not push to main; open a PR instead."

# Cross-event: from chenhg5/cc-connect
rule tests-before-commit:
  block exec "git" "commit"
    if AGENT unless after exec "go" since write "**/*.go"
  because "Run `go test ./...` before committing."
\end{verbatim}
\caption{Two \sys{} rules drawn from real projects. The first is a
per-event rule (single \texttt{execve} match); the second is a
cross-event rule requiring temporal state (\texttt{after\ldots since}).}
\Description{Code listing of two \sys{} rules from real projects.}
\label{fig:policy-example}
\end{figure}

Each clause declares its own effect that determines how the kernel
responds to a match.

\textbf{Notify}: the kernel emits a match event and allows the operation
to proceed.

\textbf{Block}: a BPF-LSM hook returns \texttt{-EPERM} before the operation
commits. Block is available only when the \texttt{bpf} LSM is active in the
running kernel; it is not silently downgraded.

\textbf{Kill}: the kernel sends \texttt{SIGKILL} to the matching task. From
a BPF-LSM hook, kill also returns \texttt{-EPERM}; from a tracepoint, kill
is post-operation termination.

The composition of multiple matching rule effects is defined by the
layered policy model above.

\subsubsection{Policy Language Scope}

The DSL is not a general-purpose mandatory access control policy. Its
primitives map to recurring agent workflow patterns: ``after tests,''
``through a review tool,'' ``data derived from untrusted input,'' and
``read-only sub-agent.'' The grammar is small so that policy authors can
translate natural-language instructions into source/sink/gate shapes without
security-policy expertise.

The intermediate DSL is a deliberate design choice. The IFC engine's
backend configuration (\texttt{struct taint\_config}) consists of hundreds
of fixed-size rule entries, bitmask allocations, and match-kind encodings
across several thousand lines of kernel code. Directly generating this
backend representation from natural language (NL$\to$backend) would require
the translating agent to understand the full engine ABI---an unrealistic
expectation. The DSL compresses each rule into a single declarative
statement that the compiler lowers into the correct binary layout,
reducing the translation task to NL$\to$DSL (one line per rule).

\subsection{System Architecture}

\subsubsection{Compiler and Kernel ABI}

The Rust compiler parses the DSL and lowers it into a C-compatible
\texttt{struct taint\_config}. Lowering allocates label bits, expands Boolean
expressions into required and forbidden bitmasks, lowers glob patterns into
exact, prefix, suffix, or wildcard match kinds, and compiles IPv4 patterns
into network/mask pairs. Human-readable rule names and reason strings
(including remediation guidance) remain in Rust metadata; the kernel receives
only fixed-size rule tables and emits integer rule identifiers.

This split keeps the eBPF program verifier-friendly. The kernel program
iterates over fixed-size arrays in read-only data using bounded loops, copies
each rule entry into a stack-local variable before matching, and uses explicit
index bounds checks for argument scanning and pattern comparison.

\subsubsection{Kernel State}

The eBPF backend maintains five BPF hash maps:

\begin{itemize}
\item \texttt{ts\_proc}: pid $\to$ label bitmask and lineage-gate state.
\item \texttt{ts\_root}, \texttt{ts\_sess}: root-process and session-level
temporal (after) gate state.
\item \texttt{ts\_file}: path hash $\to$ label bitmask.
\item \texttt{ts\_endp}: IPv4 address $\to$ label bitmask.
\end{itemize}

BPF map entries are keyed by tid (one entry per Linux thread) for
\texttt{ts\_proc} and by path hash or IPv4 address for object maps.
Updates use atomic BPF helper operations; concurrent reads from different
CPUs observe a consistent label bitmask per entry.

Tracepoint programs handle process lifecycle events (fork, exec, exit) and
post-operation file and connect events. BPF-LSM programs attach to
\texttt{bprm\_check\_security}, \texttt{file\_open}, and
\texttt{socket\_connect} for pre-operation enforcement. When BPF-LSM is not
active, the loader uses tracepoint mode and disables block-effect rules.
In tracepoint mode, hooks fire after the operation commits; a
\texttt{connect} tracepoint observes an already-established connection.
This is why \texttt{block} requires BPF-LSM: only LSM hooks
provide a pre-operation verdict. Tracepoint mode supports \texttt{notify}
(post-operation observation) and \texttt{kill} (post-operation task
termination).

The current implementation supports up to 32 sources, 32 rules, 16
transforms, and 16 gates per policy. Rule evaluation uses
\texttt{bpf\_loop} for bounded iteration, which the eBPF verifier accepts
for policies exceeding 100 rules.

\subsubsection{Unified Event Handling}

Each hook normalizes kernel arguments into a common event structure:
subject pid, object kind, access kind, backend mode, target string,
optional argument vector, and optional IPv4 address. A shared handler
resolves the subject's label bitmask, applies source labels for
read/connect events, evaluates all supported rules, emits at most one
\texttt{TAINT\_VIOLATION} event for the highest-priority matching rule, and
then applies propagation updates.

\subsection{Corrective Feedback}

The kernel match event contains the rule identifier, effect, target
string, pid, and flags indicating whether the operation was blocked or the
task was killed. The Rust runner maps the rule identifier to the
corresponding rule name and reason string, then appends a
structured payload to \texttt{.actplane/last-violation.txt}.

\begin{figure}[t]
\small
\begin{verbatim}
[ActPlane] policy rule matched
rule: tests-before-commit
effect: block
event: exec git commit
reason: Run `go test ./...` before committing.
recovery: run `go test ./...`, then retry commit.
\end{verbatim}
\caption{Example corrective-feedback payload appended after a kernel
rule match. The hook relays this payload into the next agent turn.}
\Description{Text example of an \sys{} feedback payload with rule,
effect, event, reason, and recovery fields.}
\label{fig:feedback-payload}
\end{figure}

Agent integration uses a post-tool hook. The \texttt{actplane feedback-hook}
subcommand reads new bytes from the feedback file and returns them as
\texttt{additionalContext} to compatible agent
runtimes~\cite{agentsight}. The hook is a read-only relay: it does not
re-evaluate policy in userspace. The kernel is the sole authority for rule
matches; userspace formats the result so the agent can select an alternative
execution path. A bare \texttt{-EPERM} tells the agent only that an
operation failed. The feedback payload instead identifies the declared rule
that fired and the compliant recovery path, which lets a cooperative agent
repair the plan rather than retrying the same operation.


\section{Implementation}
\label{sec:implementation}

\sys{} is implemented as a Rust driver and an eBPF program pair. The
system builds a single \texttt{actplane} binary that embeds the compiled
eBPF loader, extracts it at runtime, compiles the project's policy DSL,
and runs a target command under the observation and enforcement harness.

\subsection{Runner and Policy Discovery}

The command \texttt{actplane run -- <cmd>} discovers
\texttt{actplane.yaml} by searching upward from the working directory.
The policy YAML contains a \texttt{policy:~|} DSL block. Before the
target process begins executing, the loader seeds the stopped target's pid
with the \texttt{AGENT} label so that all descendants inherit the agent
identity. The runner prepares the feedback file
(\texttt{.actplane/last-violation.txt}) and a hook state file, exports
their paths to the child environment, waits for the loader to report
readiness, then resumes the target. This sequencing ensures that the
first target operation is observed.

\subsection{BPF Backends}

The eBPF program attaches to process lifecycle tracepoints (fork, exec,
exit), syscall tracepoints for file and connect events, and BPF-LSM hooks
for pre-operation enforcement. At load time, the loader checks whether the
running kernel has the \texttt{bpf} LSM active in
\texttt{/sys/kernel/security/lsm}. If not, it falls back to tracepoint
mode, in which the supported effects are \texttt{notify} and \texttt{kill};
\texttt{block} rules are disabled because there is no pre-operation verdict
path.

\subsection{Limitations}

\textbf{Path identity.} File labels are keyed by FNV-1a path hash rather
than inode/device pairs. Hard links and mount binds can cause the same
file to carry different labels under different paths.

\textbf{Endpoint identity.} Connect matching uses numeric IPv4 addresses.
Hostname, DNS, SNI, and HTTP-layer matching require userspace resolution or
additional instrumentation.

\textbf{Argument predicates.} Positional argument predicates inspect
arguments after exec via the tracepoint argument buffer. Pre-operation
blocking of argument predicates requires an argument cache before
\texttt{bprm\_check\_security}, which is not yet implemented.

\textbf{Label precision.} Labels propagate at object granularity. A process
that reads one byte from a labeled file acquires the full label set. This
can produce over-tainting relative to byte-level
systems~\cite{kemerlis2012libdft}. Section~\ref{sec:evaluation} measures the
false-positive rate and validates that declassification and gate paths
prevent unnecessary harness intervention.
