\section{Evaluation Plan}
\label{sec:evaluation}

The evaluation is organized around the following reproducible experiments.
Result numbers are not filled in until the corresponding experiment is run with
fixed agent versions, logs, and statistical analysis.

\subsection{Research Questions}

The evaluation targets six questions.

\begin{description}
\item[RQ1: Bypass coverage.] Does OS-level enforcement catch the same policy
violation through direct tool calls, shell strings, subprocesses, SDK paths, and
direct syscalls?
\item[RQ2: Corrective feedback.] Does semantic feedback improve task completion
and reduce repeated violations compared with failure without a reason?
\item[RQ3: False positives.] Do benign tasks and sanctioned gate/declassify
paths proceed without unnecessary failure?
\item[RQ4: Overhead.] What are the per-event and end-to-end costs of label
propagation and rule checks?
\item[RQ5: Real-policy coverage.] How many real instructions from the corpus
are observable, expressible, robust, and temptable?
\item[RQ6: Baselines.] Which capabilities are unique relative to tool-layer
agent guardrails and kernel provenance/enforcement systems?
\end{description}

\subsection{Experiments}

\begin{table*}[t]
\centering
\caption{Planned evaluation matrix. Result fields remain \textsc{TBD}.}
\Description{A table listing planned experiments and planned result fields.}
\begin{tabular}{p{0.12\textwidth}p{0.26\textwidth}p{0.26\textwidth}p{0.24\textwidth}}
\toprule
Experiment & Method & Primary metrics & Result status \\
\midrule
E1: Cross-path coverage &
Run the same forbidden action through tool call, nested shell, subprocess, SDK/helper, and direct syscall triggers. Compare against tool-layer baselines. &
Coverage matrix over policy by path; feedback delivery. &
\textsc{TBD} \\
\midrule
E2: Feedback loop &
Run Codex and Claude-style agents under prompt-only, audit, fail-without-reason, and fail-with-feedback conditions. &
Task completion, repeated violations, repair iterations, token/turn cost. &
\textsc{TBD} \\
\midrule
E3: False positives &
Run allowed tasks and sanctioned paths: redaction, test-before-commit, confirmation, and lineage mediation. &
False positive rate, allowed-path success. &
\textsc{TBD} \\
\midrule
E4: Overhead &
Microbench fork/exec/read/write/connect and macrobench real agent tasks with and without ActPlane. &
p50/p99 latency, wall time, memory/map usage, rule-table scaling. &
\textsc{TBD} \\
\midrule
E5: Corpus coverage &
Double-code real instruction-file rules with D1--D7, report agreement, and derive policies/tasks mechanically. &
Observable, expressible, robust, temptable fractions; Cohen's kappa. &
\textsc{TBD} \\
\midrule
E6: Baseline comparison &
Compare against CamQuery, Tetragon, Progent, AgentSpec, and built-in agent sandboxes where runnable. &
Layer, enforce/detect, cross-channel taint, bypass resistance, feedback. &
\textsc{TBD} \\
\bottomrule
\end{tabular}
\label{tab:eval-plan}
\end{table*}

\subsection{Feedback Conditions}

The agent-facing experiment uses four conditions to isolate the contribution of
semantic feedback.

\begin{itemize}
\item C1: prompt-only. The natural-language rule is present, but there is no
ActPlane enforcement.
\item C2: audit. ActPlane observes and reports, but does not fail the action.
\item C3: fail without ActPlane reason. The violating action fails through
\texttt{-EPERM} or \texttt{SIGKILL}, but the agent receives no structured
remediation.
\item C4: fail with feedback. The action fails and ActPlane injects the rule
reason and remediation through the feedback file and hook bridge.
\end{itemize}

The main comparison is C3 to C4: the net benefit of the corrective feedback
loop. C1 to C2 estimates how often the prompt-only harness fails; C2 to C3
separates observation from action failure.

\subsection{Artifact Expectations}

The artifact should include policies, trigger scripts, agent prompts, raw logs,
statistical notebooks, and a container or VM description pinning kernel, BPF,
agent, and model versions. The final paper should avoid reporting any number
that cannot be regenerated from these artifacts.
