\documentclass[sigplan,10pt]{acmart}

\setcopyright{none}
\renewcommand\footnotetextcopyrightpermission[1]{}
\settopmatter{printfolios=true}
\pagestyle{plain}

\acmConference[ATC '26]{2026 ACM SIGOPS Annual Technical Conference}{November 15--18, 2026}{Hyatt Hotel, Shatin, Hong Kong}
\acmYear{2026}
\acmISBN{}
\acmDOI{}

\title[ActPlane]{ActPlane: OS-Enforced Agent Harnesses}

\author{Paper \#XXX}
\affiliation{%
  \institution{Anonymous Submission}
  \city{}
  \country{}}

\begin{abstract}
AI coding agents are guided by natural-language instruction files that express
behavioral contracts: do not leak secrets, run tests before committing, use
only sanctioned tools. These contracts are declared at the level of
\emph{intent}---what the agent should and should not do---but their effects
occur as \emph{behavior}---actual operations such as file access, process
execution, and network connections. Between the two sits the agent's
\emph{action} layer: tool calls whose mapping to OS behavior is many-to-many
and opaque.  A single tool invocation can trigger arbitrary system calls, and the
same OS operation can be reached through different tool paths. 
Existing enforcement operates at a single level: action-level
guards check tool calls but miss effects that reach the OS through subprocesses
or direct system interfaces; behavior-level enforcement (syscall filters,
sandboxes) is unbypassable but returns opaque failures disconnected from
intent. 

Today's agent harnesses---the non-model infrastructure around an
agent to enforce the behavioral contracts---keep their  components entirely at the
application layer, leaving this \emph{intent--behavior gap} unbridged.
ActPlane is the OS-level enforcement component of an agent harness, allowing agent
and human. declares behavioral contracts in a compact DSL and ActPlane compiles them 
into labeled information-flow rules enforced by an eBPF/BPF-LSM backend. 
Named labels propagate across process, file, and network objects at kernel hooks, 
tracking which sources have
influenced which subjects. When a violation is detected, the kernel reports it
and ActPlane returns corrective feedback to the agent: which contract was
violated, why, and what alternative path satisfies it. The agent receives not
\texttt{Permission denied} but a reminder of its own declared intent.we evaluates
ActPlane on bypass coverage across tool paths, agent recovery rate 
under feedback conditions, false positives, and per-event overhead.
\end{abstract}

\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded and cyber-physical systems</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10002978.10003022.10003023</concept_id>
  <concept_desc>Security and privacy~Software and application security</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10011007.10010940.10010941.10010949</concept_id>
  <concept_desc>Software and its engineering~Software safety</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[300]{Computer systems organization~Embedded and cyber-physical systems}
\ccsdesc[300]{Security and privacy~Software and application security}
\ccsdesc[300]{Software and its engineering~Software safety}

\keywords{AI agents, eBPF, BPF-LSM, labeled information-flow control, information-flow control, provenance, corrective feedback}

\begin{document}
\maketitle

\input{sections/01-introduction}
\input{sections/02-motivation}
\input{sections/03-model}
\input{sections/04-design}
\input{sections/05-implementation}
\input{sections/06-evaluation}
\input{sections/07-related}
\input{sections/08-discussion}

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
